\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-20 doc/paper-jss/sdmTMB-paper.Rnw
Edits

In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally
expensive and methods of working directly with the sparse precision matrix
($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are much more efficient. We introduce
the SPDE approach to working with the precision matrix in the software section,
but for now, note that we can instead represent our spatial random field as
\[
\boldsymbol{\omega} \sim
  \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
\]
where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
%
% The SPDE approach does just that, ... have been developed lets us stay with precision and so as precision:
%
% The approach involves finding an SPDE w

@@

\subsection{Delta models}

@@

\section{Software}

@@

Importantly, \texttt{dependencies = TRUE} includes the \proglang{R}-INLA
package \citep{rue2009, lindgren2011, lindgren2015}, which is not on CRAN but is
needed for the \texttt{make\_mesh()} function below. Soon, we intend to switch
to using the \pkg{fmesher} \citep{fmesher} package to eliminate this dependency.

@@

We begin with a simple species distribution model of encounter probability of
Pacific Cod (\emph{Gadus macrocephalus}) from a trawl survey conducted in Queen
Charlotte Sound, British Columbia, Canada. The purpose of our example is to
illustrate the need for spatial random fields. This survey is conducted by
Fisheries and Oceans Canada (REF) and follows a depth-stratified random sampling
design, resulting in a georeferenced dataset. The data frame \texttt{pcod} is
available as package data in \pkg{sdmTMB}. Relevant columns include latitude,
longitude, Universal Transverse Mercator (UTM) coordinates, bottom depth, and
encounter (\texttt{present = 1}) vs.~non-encounter (\texttt{present = 0}) of
Pacific Cod for a given survey sample.

@@

\sloppy We then create a mesh object that contains triangulation and projection
matrices needed to apply the SPDE approach using \texttt{make\_mesh()}.
\texttt{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher}
package. The argument \texttt{cutoff} defines the minimum allowed distance
between mesh vertices in the units of \texttt{X} and \texttt{Y} (km). We can
also specify additional arguments, in this case passed to
\texttt{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 20 km
and 40 km for the inner and outer mesh, and an offset border width of 5 km and
40 km for the inner and outer meshes. Alternatively, we could have created any
mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supply it to the
\texttt{mesh} argument in \texttt{make\_mesh()}. We can inspect mesh object with
the associated plotting method (Figure~\ref{fig:dog-binomial-mesh2}). Our mesh
has 238 (\texttt{mesh\_pcod\$mesh\$n}) vertices. Guidelines for what makes
a good mesh are described in TODO.

@@

<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
mesh_pcod2 <- make_mesh(
  pcod, c("X", "Y"),
  fmesher_func = fmesher::fm_mesh_2d_inla,
  cutoff = 8,
  convex = 10,
  max.edge = c(20, 40),
  offset = c(5, 40)
)
plot(mesh_pcod2)
@

@@

By setting \texttt{effects = "ran\_pars"}, \texttt{tidy()} will return random
field parameters, where \texttt{sigma\_O} is \(\sigma_O\) in the above
equations---the standard deviation of the spatial random field \(\bm{\omega}\).

@@

Our model can be written as TODO

\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t}
\right),
\end{aligned}
\]

where \(\bm{\beta}\) are coefficients associated with the main effects,
\(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\)
represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the spatial
field (equivalent to a spatial intercept) and \(\epsilon_{\bm{s},t}\) represents
spatiotemporal fields.

The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1)
process,

\[
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right),
\end{aligned}
\]

where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\)
at time $t-1$ and $t$.

@@

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta or hurdle distribution: delta-gamma and
delta-lognormal \citep{aitchison1955}, and delta-Poisson-link-gamma and
delta-Poisson-link-lognormal \citep{thorson2018poisson}. The Poisson-link models
depart from conventional delta models in treating the probability of occurrence
at a given location ($p_i$) as a Poisson process rather than a Bernoulli one,
$p_i = 1 - \exp(a_i \cdot n_i)$, where $n_i$ is the total number of individuals
available to be encountered, and $a_i$ is a measure of sampling effort
\citep{thorson2018poisson}. Note that we choose to omit the spatiotemporal
random fields from the binomial model component (\texttt{spatiotemporal
= list("off", "iid")}) because initial fits indicated this random field was
collapsing to zero indicating minimal changes in encounter spatial distribution
from year to year.

>>>>

In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally
expensive and methods of working directly with the sparse precision matrix
($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are much more efficient. We introduce
the SPDE approach to working with the precision matrix in section \ref{sec:software},
but for now, note that we can instead represent our spatial random field as
\[
\boldsymbol{\omega} \sim
  \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
\]
where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
%
% The SPDE approach does just that, ... have been developed lets us stay with precision and so as precision:
%
% The approach involves finding an SPDE w

@@

\subsection{Delta models}\label{sec:delta}

@@

\section{Software}\label{sec:software}

@@

Importantly, \texttt{dependencies = TRUE} includes the \proglang{R}-INLA
package \citep{rue2009, lindgren2011, lindgren2015}, which is not on CRAN but is
required to generate matrices for the SPDE.
%Soon, we intend to switch to using the \pkg{fmesher} \citep{fmesher} package to eliminate this dependency.

@@

We begin with a simple species distribution model of encounter probability of
Pacific Cod (\emph{Gadus macrocephalus}) from a trawl survey conducted in Queen
Charlotte Sound, British Columbia, Canada. The purpose of our example is to
illustrate the need for spatial random fields. This survey is conducted by
Fisheries and Oceans Canada and follows a depth-stratified random sampling
design, resulting in a georeferenced dataset. The data frame \texttt{pcod} is
available as package data in \pkg{sdmTMB}. Relevant columns include latitude,
longitude, Universal Transverse Mercator (UTM) coordinates, bottom depth, and
encounter (\texttt{present = 1}) vs.~non-encounter (\texttt{present = 0}) of
Pacific Cod for a given survey sample.

@@

\sloppy We then create a mesh object that contains triangulation and projection
matrices needed to apply the SPDE approach using \texttt{make\_mesh()}.
\texttt{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher}
package. The argument \texttt{cutoff} defines the minimum allowed distance
between mesh vertices in the units of \texttt{X} and \texttt{Y} (km). We could create a basic mesh just specifying this:

@@

We can also specify additional arguments, in this case passed to \texttt{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders. Alternatively, we could have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supplied it to the \texttt{mesh} argument in \texttt{make\_mesh()}. For more irregularly shaped areas, we could have specified a non-convex hull with \texttt{convex} and \texttt{concave} arguments.
Recommendations... The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).

We can visualize the mesh object with the associated plotting method (Figure~\ref{fig:dog-binomial-mesh2}).
Our mesh has 563 (\texttt{mesh\_pcod2\$mesh\$n}) vertices.
Mesh complexity has the single largest influence on the speed of fitting these models...

% https://haakonbakkagit.github.io/btopic104.html
% https://becarioprecario.bitbucket.io/spde-gitbook/index.html

<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
mesh_pcod2 <- make_mesh(
  pcod, c("X", "Y"),
  fmesher_func = fmesher::fm_mesh_2d_inla,
  cutoff = 8,
  max.edge = c(10, 40),
  offset = c(10, 40)
)
plot(mesh_pcod2)
@

@@

By setting \texttt{effects = "ran\_pars"}, \texttt{tidy()} will return random
field parameters, where \texttt{sigma\_O} is the marginal standard deviation of the spatial random field \(\bm{\omega}\) (`O' for `Omega').

@@

Our model can be written as

\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t}
\right),
\end{aligned}
\]
where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the spatial field (equivalent to a spatial intercept) and \(\epsilon_{\bm{s},t}\) represents spatiotemporal fields.

The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1)
process,
\[
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right),
\end{aligned}
\]
where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\)
at time $t-1$ and $t$.

@@

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.
% The Poisson-link models depart from conventional delta models in treating the probability of occurrence at a given location ($p_i$) as a Poisson process rather than a Bernoulli one, $p_i = 1 - \exp(a_i \cdot n_i)$, where $n_i$ is the total number of individuals available to be encountered, and $a_i$ is a measure of sampling effort \citep{thorson2018poisson}.
Note that we choose to omit the spatiotemporal random fields from the binomial model component (\texttt{spatiotemporal = list("off", "iid")}) because initial fits indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.
